\chapter{Conclusion and Recommendations}

\section{Conclusion}

This research developed and evaluated a multimodal framework for Automatic Personality Recognition (APR) of Filipino Instagram users using the PagkataoKo dataset. The study pursued six objectives: finalizing a suitable data subset, identifying relevant data characteristics, extracting multimodal features, building prediction models, identifying indicative features per trait, and evaluating performance against unimodal approaches.

The feature extraction process generated three modality streams: visual features from ResNet50, HSV color analysis, and Imagga object tags; textual features from TF-IDF and Word2Vec; and behavioral metadata from post and following counts. These were combined through early fusion and used to train Logistic Regression, SVM, and XGBoost classifiers for each Big Five trait.

The results reveal that optimal feature sets vary across personality traits:

\begin{itemize}
	
	\item \textbf{Openness} was best predicted by visual features alone (\(\kappa = 0.2762\)), specifically ResNet50 and HSV.
	
	\item \textbf{Conscientiousness} benefited from a mixed-feature approach (\(\kappa = 0.2451\)), combining TF-IDF, Word2Vec, and metadata with SVM. This suggests that organized behavior is expressed through both specific word choices and posting regularity.
	
	\item \textbf{Extraversion} achieved the strongest performance (\(\kappa = 0.3684\)) using Word2Vec and HSV features with Logistic Regression.
	
	\item \textbf{Agreeableness} was best captured by Word2Vec, HSV, and metadata with SVM (\(\kappa = 0.2129\)).
	
	\item \textbf{Neuroticism} proved most challenging, with unimodal TF-IDF achieving the highest performance (\(\kappa = 0.1615\)).
	
\end{itemize}

Only the personality traits Openness, Conscientiousness, and Extraversion garnered a fair kappa score (0.21 and above). Agreeableness and Neuroticism proved difficult.

Several key findings emerge from this work. First, predictive performance is inherently trait-dependent: no single modality or feature type dominates across all five dimensions. Second, while multimodal fusion generally enhances robustness, certain traits—particularly Neuroticism and Openness—are most effectively captured through specific unimodal features; TF-IDF for Neuroticism, and Resnet50 and HSV features for Openness. Third, metadata features, though only being two features, improved the performance for Conscientiousness and Agreeableness. Conscientiousness had a + 0.0491 kappa improvement, while Agreeableness had a +.0199 kappa improvement. Fourth, the Imagga object tags yielded negligible predictive value across all traits when processed with TF-IDF weighting. The 30\% confidence filter completely eliminated predictive signal, producing zero kappa across all models.

These findings advance Filipino APR by demonstrating that mixed-feature approaches—combining text, image, and metadata features— can improve predictive performance. The work also establishes that Instagram's multimodal nature provides complementary personality signals that text-only Twitter-based studies cannot capture.
\section{Recommendations}
\label{sec:recommendations}

Future research should explore using multimodal features as well. While this study identified that some traits work better unimodal (Openness and Neuroticism), others perform better when multimodal is used. Regarding model architectures, the findings suggest a transition toward more complex systems. While the current dataset size might make standard Multilayer Perceptrons (MLPs) less effective, the use of multimodal fusion networks can be further utilized for capturing non-linear relationships between image and text data.

Specific technical improvements can also be made to the feature extraction process. For text features, exploring Convolutional Neural Networks (CNNs) could capture local context more effectively and push the performance of Word2Vec embeddings further. Additionally, since the domain of social media data often differs from standard training sets, exploring other pre-trained transformers could bridge this gap. For image features, while ResNet provided strong results, the Imagga tags yielded negligible ones. To our knowledge, this study represents the first attempt to apply TF-IDF weighting to Imagga object tags in an APR context, and our results suggest this approach is ill-suited for tag-based features. Future work should explore alternative methods:

\begin{itemize}
	\item \textbf{Raw tag frequencies or binary indicators} without IDF weighting, as the discriminative value of object tags may lie in simple presence/absence patterns rather than document-frequency rarity.
	\item \textbf{Tag embeddings} that capture semantic relationships between detected objects, similar to word embeddings, rather than treating tags as independent terms.
	\item Alternative \textbf{object detection models} (e.g., YOLO, Detectron2) that allow greater control over the tag vocabulary and confidence thresholding, unlike the private Imagga API.
	\item Different \textbf{confidence thresholds} that preserve sufficient tag coverage while reducing noise, as our 30\% filter proved too aggressive.
\end{itemize}

Additionally, an ensemble approach combining various models for prediction could be explored in future research. While this research simply selected the best-performing combination regardless of the model used, the results in Table~\ref{tab:algorithm_gaps} show that other models do not fall too far behind in performance and may therefore still be used more effectively. The data shows that the traits of Extraversion, Agreeableness, and Neuroticism have second-best models that are close in performance but use a different algorithm than the best model. While the Extraversion and Agreeableness models’ best and second-best feature sets are shown to be somewhat similar, as both models in both traits contain the Word2Vec feature, Neuroticism’s is interesting in that the best and second-best models use completely different features: TF-IDF in the best model using SVM, and Metadata in the second-best model using XGBoost.

Future work should also employ k-fold cross-validation to obtain a more general and stable performance to better detect overfitting, which is observed across all models/combinations.

While this study covered images, only images from user posts were considered, and profile pictures from the dataset were left untouched. Even platforms that do not primarily focus on images, such as Twitter, may make use of the visual information contained in profile pictures. As some traits find predictive value in image features, considering these as well could be further explored in future studies. 

Table~\ref{tab:descriptive_stats} presented the finding that for Conscientiousness and Extraversion, correctly classified users had more posts on average, with the gap being especially noticeable on Extraversion, having a mean of 213 and 143 for correctly and incorrectly classified users, respectively. With this in mind, it may be concluded that having more posts gives the models more valuable information to work with, thereby increasing model performance. Given that this study used a 300-post cap per user, it may be beneficial to explore using all data without the cap in future studies, if within resource constraints.

